---
title: 'Lousville Dockless Vehicles: Exploratory Data Analysis'
author: 'Luca Baggi'
---

# Load packages
```{r, message=FALSE}
library(tidyverse)

# for enhancing ggplot
library(ggridges)
library(ggthemes)

# set a theme to stay for the rest of the plotting
theme_set(theme_fivethirtyeight())
```

# Load data

```{r}
url = 'https://raw.githubusercontent.com/baggiponte/escooters-louisville-python/main/data/escooters_od.csv'

trips <-
  read_csv(
    url,
    col_types = cols(
      StartTime = col_datetime(format = '%Y-%m-%d %H:%M:%S'),
      EndTime = col_datetime(format = '%Y-%m-%d %H:%M:%S')
    )
  ) %>%
  mutate(
    Duration = as.integer(Duration),
    # convert in meters
    Distance = as.integer(round(Distance * 1609)),
    # turn into factors
    StartNH = as.factor(StartNH),
    EndNH = as.factor(EndNH),
    # rename TripID
    TripID = 1:length(TripID),
    # add a dummy for short trip (with short duration and distance)
    ShortTrip = as.factor(ifelse((Duration <= 1 | Distance <= 100), 1, 0)),
    # first covid death reported is on March 21st, 2020
    Covid = as.factor(ifelse(StartTime > '2020-03-20 23:59:59', 1, 0))
  ) %>%
  as_tibble()
```

Two features have been added:

* `Covid` indicates if the trip happened during the COVID-19 emergency.
* `ShortTrip` indicates if the trip is actually a very short one: shorter than 1 minute, with less than 100 meters run.

```{r, echo=FALSE}
trips %>%
  count(ShortTrip) %>%
  mutate(count = n/sum(n))
```

Less than 1 trip on 5 is very, very short. This may actually be a really low threshold:

```{r, echo=FALSE}
trips %>%
  count(Covid) %>%
  mutate(count = n/sum(n))
```
Only some 25.000 trips happened during COVID-19 emergency.

# Boxplot

## Boxplot as it is

To obtain a first impression, we can draw the boxplots of `Duration`:

```{r, echo=FALSE}
trips %>%
  ggplot(aes(x = Duration)) +
  geom_boxplot() +
  ggtitle('Duration Boxplot')
```

And of `Distance`:

```{r, echo=FALSE}
trips %>%
  ggplot(aes(x = Distance)) +
  geom_boxplot() +
  ggtitle('Distance Boxplot')
```


There are way to many outliers: we can't see anything!

## Create a Filter

Note that there are some values of distance which are negative: we should also deal with them.

```{r, echo=FALSE}
trips_filter <-
  trips %>%
  filter(
    # Duration between 0 and 30 minutes
    Duration > 0 & Duration <= 30 &
    # Distance between 0 and 5km
    Distance > 0 & Distance <= 5000
  )
```

We choose to create a `trips_filter` dataset with `Duration` smaller than 30 minutes and `Distance` smaller than 5km.

```{r}
trips_filter %>%
  ggplot(aes(x = Duration)) +
  geom_boxplot() +
  ggtitle('Boxplot of Duration')

trips_filter %>%
  ggplot(aes(x = Duration, y = StartNH)) +
  geom_boxplot() +
  ggtitle('Duration, by Starting Neighborhood')

trips_filter %>%
  ggplot(aes(x = Distance)) +
  geom_boxplot() +
  ggtitle('Boxplot of Distance')

trips_filter %>%
  ggplot(aes(x = Distance, y = StartNH)) +
  geom_boxplot() +
  ggtitle('Distance, by Starting Neighborhood')
```

Filtering is almost always arbitrary. Here we are not conservative at all especially because we are left nonetheless with more than 350.000 observations - a real challenge for our hardware to process!

However, there are theoretical reasons behind this nonetheless: trips lasting longer than half an hour seem suspicious - we would need idle time to make a more thorough analysis. The `distance` threshold is not varying very much: only some 10.000 observations get eliminated. This will result in cleaner histograms.

This will indeed underrepresent trips from the outermost neighbourhoods, but we would delete most of these regardless.

# Histograms for Distance and Duration

Getting a preliminary idea of the distributions is a good practice which allows to narrow down the interesting patterns:

```{r}
trips_filter %>%
  ggplot(aes(x = Distance)) +
  geom_histogram(fill = 'darkseagreen', bins = 60) +
  ggtitle('Distance Histogram')

trips_filter %>%
  ggplot(aes(x = Duration)) +
  geom_histogram(fill = 'cornflowerblue', bins = 30) +
  ggtitle('Duration Histogram')
```

But on their own these graphs aren't very informative. We might want to see if COVID-19 has any effects:

```{r}
trips_filter %>%
  ggplot(aes(x = Distance, fill = Covid)) +
  geom_histogram(bins = 60, alpha = 0.6)

trips_filter %>%
  ggplot(aes(x = Duration, fill = Covid)) +
  geom_histogram(bins = 60, alpha = 0.6)
```

It turns out, this does not tell very much, too!

# Ridges, Histogram, Densities

## COVID-19 Effects

With `ggridges`, we can plot these:

```{r}
trips_filter %>%
  ggplot(aes(x = Distance, y = Covid)) +
  geom_density_ridges() +
  ggtitle('Distance: effects of COVID-19') +
  ylab('0: no COVID-19')

trips_filter %>%
  ggplot(aes(x = Duration, y = Covid)) +
  geom_density_ridges() +
  ggtitle('Duration: effects of COVID-19') +
  ylab('0: no COVID-19')
```

Clearly COVID-19 shifts escooters usage. We can formulate a couple of hypotheses, which we won't be statistically testing:

1. COVID-19 reduces the trip distance.
2. COVID-19 however seems to have increased the variance of the trip duration.

```{r}
trips_filter %>%
  filter(Covid == 0) %>%
  ggplot(aes(Duration, fill = ShortTrip)) +
  geom_histogram(alpha = 0.8) +
  ggtitle('Duration without COVID-19')

trips_filter %>%
  filter(Covid == 1) %>%
  ggplot(aes(Duration, fill = ShortTrip)) +
  geom_histogram(alpha = 0.8) +
  ggtitle('Duration with COVID-19')
```

```{r}
trips_filter %>%
  filter(Covid == 0) %>%
  ggplot(aes(Distance, fill = ShortTrip)) +
  geom_histogram() +
  ggtitle('Distance without COVID-19')

trips_filter %>%
  filter(Covid == 1) %>%
  ggplot(aes(Distance, fill = ShortTrip)) +
  geom_histogram() +
  ggtitle('Distance with COVID-19')
```

Quite the contrary, it seems that the duration of longer trips is the only feature to have diminished. To really test that, we would need to control for the time period where the trips actually happened!

# Ridges and Starting Neighbourhoods

Using `ggridges`, we can plot some smoothed histograms to get the idea of the distributions for each starting neighbourhood:

```{r}
trips_filter %>%
  ggplot(aes(x = Distance, y = StartNH)) +
  geom_density_ridges() +
  ggtitle('Distance by starting neighbourhood') +
  ylab('0: no COVID-19')

trips_filter %>%
  ggplot(aes(x = Duration, y = StartNH)) +
  geom_density_ridges() +
  ggtitle('Duration by starting neighbourhood') +
  ylab('0: no COVID-19')
```