---
title: "Louisville Dockless Vehicles: Advanced Visualisation"
author: 'Luca Baggi'
---

# Load Packages

Now we want to create some visualisations of the flows from `StartNH` to `EndNH`. We will use the following libraries:

```{r, message=FALSE}
library(tidyverse)

# alluvial and sankey plots
library(ggalluvial)

# chord plots
library(circlize)
```

# Load Data

From before, we did:

```{r}
url = 'https://raw.githubusercontent.com/baggiponte/escooters-louisville-python/main/data/escooters_od.csv'

trips <-
  # read the data
  read_csv(url, col_types = cols(
      StartTime = col_datetime(format = '%Y-%m-%d %H:%M:%S'),
      EndTime = col_datetime(format = '%Y-%m-%d %H:%M:%S')
    )
  ) %>%
  # manipulate cols:
  mutate(
    Duration = as.integer(Duration),
    # convert in meters
    Distance = as.integer(round(Distance * 1609)),
    # turn into factors
    StartNH = as.factor(StartNH),
    EndNH = as.factor(EndNH),
    # recode TripID
    TripID = 1:length(TripID),
    # add a dummy for short trip (with short duration and distance)
    ShortTrip = as.factor(ifelse((Duration <= 1 | Distance <= 100), 1, 0)),
    # first covid death reported is on March 21st, 2020
    Covid = as.factor(ifelse(StartTime > '2020-03-20 23:59:59', 1, 0))
  ) %>%
  # remove the outliers
  filter(
    # Duration between 0 and 30 minutes
    Duration > 0 & Duration <= 30 &
    # Distance between 0 and 5km
    Distance > 0 & Distance <= 5000
  ) %>%
  as_tibble()
```

# Data Wrangling

Before plotting, we need to create the two way frequency table:

```{r}
twoway_freq <-
  trips %>%
  select(StartNH, EndNH) %>%
  table() %>%
  as_tibble() %>%
  # remove the null combinations
  filter(n != 0) %>%
  arrange(desc(n))

twoway_freq
```

# Senkey Diagram using `ggalluvial`

This already has the desired long format of origin-destination, so we can directly work with the Senkey diagram. To reduce dimensionality, we might want to filter the values.

```{r, echo=FALSE}
twoway <-
  trips %>%
  select(StartNH, EndNH) %>%
  table() %>%
  as_tibble() %>%
  arrange(desc(n)) %>%
  filter(n > 1000) %>%
  mutate(
    EndNH = paste0(EndNH, ' ')
  )

twoway %>%
  ggplot(aes(
    axis1 = StartNH,
    axis2 = EndNH,
    y = n
  )) +
  scale_x_discrete(limits = c("StartNH", "EndNH"), expand = c(.2, .05)) +
  xlab("Start and End Neighbourhood") +
  geom_alluvium(aes(fill = n)) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)))
```

# Drawing a Chord Plot

The matrix we are going to use is still the two-way frequency table (or, in proper terms, adjacency matrix).

## Define some helper functions

We define two helper functions which will improve readability. The original code is available in the GitHub repo.

```{r, echo=FALSE}
prep_cord_data <- function(freq_table) {
  freq_table %>%
  transmute(
    StartNH = as.integer(as.factor(StartNH)),
    EndNH = as.integer(as.factor(EndNH))
  ) %>%
  bind_cols(freq_table %>% select(n)) %>%
  arrange(StartNH, EndNH)
}
  
```

Then we initialise the chord diagram and set some more parameters:

```{r}
# not compatible in tidy grammar, sigh

circos.clear() # reset params

circos.par( # set a bunch of stuff
  start.degree = 90,
  gap.degree = 4,
  track.margin = c(-0.1, 0.1),
  points.overflow.warning = FALSE
)

# set plot parameters:
par(mar = # margins
      rep(0,4)) # 0 repeated 4 times
```

And then proceed with the creation of the plot. Once again, we will create a custom function. The full code is available on the GitHub [repo](https://github.com/baggiponte/escooters-louisville-r).

```{r, echo=FALSE}
# base plot

chord_plot <- function(my_data) {
  
  chordDiagram(
    x = my_data, 
    transparency = 0.25,
    directional = 1,
    direction.type = c("arrows", "diffHeight"), 
    diffHeight  = -0.04,
    annotationTrack = "grid", 
    annotationTrackHeight = c(0.05, 0.1),
    link.arr.type = "big.arrow", 
    link.sort = TRUE, 
    link.largest.ontop = TRUE)
  
  # text and axis
  circos.trackPlotRegion(
    track.index = 1, 
    bg.border = NA, 
    panel.fun = function(x, y) {
      
      xlim = get.cell.meta.data("xlim")
      sector.index = get.cell.meta.data("sector.index")
      
      # Add names to the sector
      circos.text(
        x = mean(xlim), 
        y = 3.2, 
        labels = sector.index, 
        facing = "bending", 
        cex = 0.8
        )
    }
  )
}

```

And finally we can plot the final results:

```{r}
# with all neighbourhoods
twoway_freq %>%
  prep_cord_data() %>%
  chord_plot()

# only top neighbourhoods
twoway %>%
  prep_cord_data() %>%
  chord_plot()
```